Supplementary methods

### 2020-08-25 text/manuscript.Rmd

We investigated this dataset further using multiple correspondence analysis
(MCA) [@JSSv025i01], and visually summarized the analysis with UMAP (**figure 2**). We used
the resulting UMAP graph coordinates to perform density based clustering with
HDBSCAN, which resulted in identification of 10 well-defined clusters capturing
about 80% of the tumor samples. To capture the remaining samples into one of
these 10 clusters we used kNN (see **Supplementary Methods** for details on all
clustering algorithms).

>>>>

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
2**) [@McInnes2018]. We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on all clustering algorithms).

### 2020-08-25 text/supplemental_methods.Rmd

In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information by selecting the eigenvalue with the most
explanatory power, using the average of 100 permutations of the data as baseline
(**figure SM1A**). We then chose the maximum eigenvalue for which the p-value
remained ≤ 0.05 (see cutoff in **figure SM1B**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several reproducible projections
regardless of random seed setting. A representative version of this 3D UMAP is
shown in **figure SM1C**, rotated to enhance the visibility of the major features.
Following this spatial mapping we attempted to define groupings of similar
tumors within the spatial map using HDBSCAN, which performs hierarchical
clustering and provides metrics of cluster stability and probabilities of
cluster membership for each node. However, HDBSCAN is sensitive to several
parameters; key for our analysis are the minimum number of tumor samples in a
cluster that capture the maximum number of tumor samples, measured by
probability of membership of ≥ 5% in at least one cluster. Thus, we created a
score metric as the fraction of classified tumors with max probability < 5% in
one cluster and chose a cluster size of 92 to minimize the score function
(**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure SM1E**).
This classified 6,038 out of 7,607 tumors but still resulted in a significant
fraction of unclassified tumor samples. Since we ultimately wish to be able to
classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
analysis, which computes a similarity metric to every tumor in the set and then
lets the _k_ most similar tumors "vote" as to the identity of the query tumor
sample based on their cluster labels. We set _k_ to be the square root of
the number of tumor samples (87). Using this method, we assigned cluster
membership to the remaining tumors (**figure SM1F**).

>>>>

In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information by selecting the eigenvalue with the most
explanatory power, using the average of 100 permutations of the data as baseline
(**figure SM1A**). We then chose the maximum eigenvalue for which the p-value
remained ≤ 0.05 (see cutoff in **figure SM1B**). Then we performed a UMAP
analysis [@McInnes2018], both in order to summarize the MCA graphically, and as
a preprocessing step to boost the performance of density based clustering. The
resulting map was notable for its lobed structure, with several reproducible
projections regardless of random seed setting. A representative version of this
3D UMAP is shown in **figure SM1C**, rotated to enhance the visibility of the
major features. Following this spatial mapping we attempted to define groupings
of similar tumors within the spatial map using HDBSCAN, which performs
hierarchical clustering and provides metrics of cluster stability and
probabilities of cluster membership for each node [@hdbscan]. However, HDBSCAN
is sensitive to several parameters; key for our analysis are the minimum number
of tumor samples in a cluster that capture the maximum number of tumor samples,
measured by probability of membership of ≥ 5% in at least one cluster. Thus, we
created a score metric as the fraction of classified tumors with max probability
< 5% in one cluster and chose a cluster size of 92 to minimize the score
function (**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure SM1E**).
This classified 6,038 out of 7,607 tumors but still resulted in a significant
fraction of unclassified tumor samples. Since we ultimately wish to be able to
classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
analysis, which computes a similarity metric to every tumor in the set and then
lets the _k_ most similar tumors "vote" as to the identity of the query tumor
sample based on their cluster labels. We set _k_ to be the square root of the
number of tumor samples (87). Using this method, we assigned cluster membership
to the remaining tumors (**figure SM1F**).
